---
title: "Getting started with gbtmkit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with gbtmkit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
# The estimation engine (trajeR) is an optional dependency; only run the fitting
# chunks when it is available so the vignette also builds without it.
has_trajer <- requireNamespace("trajeR", quietly = TRUE)
has_ggplot <- requireNamespace("ggplot2", quietly = TRUE)
```

`gbtmkit` turns group-based trajectory modeling (GBTM) into a reproducible
pipeline that follows the GRoLTS reporting checklist. Estimation is delegated to
interchangeable backends behind one interface, so the same workflow handles
binary and continuous outcomes, and the fit diagnostics (entropy, APPA, OCC,
group proportions) are computed the same way regardless of engine.

```{r setup}
library(gbtmkit)
```

## The data

The package ships two entirely synthetic datasets with known ground-truth
groups. `sim_binary` has a binary outcome measured on ten occasions, with four
latent trajectory shapes of mixed polynomial order: a linear rising group, a
cubic rise-peak-decline group, a cubic decline-trough-recovery group, and a
linear falling group:

```{r}
data("sim_binary", package = "gbtmkit")
head(sim_binary)
```

## 1. Describe the model with a spec

`gbtm_spec()` records *what* to model -- the outcome and time columns (by name),
the id, and the outcome family -- and validates it, independent of which engine
will fit it.

```{r}
spec <- gbtm_spec(
  sim_binary,
  outcomes = paste0("y", 1:10),
  time     = paste0("t", 1:10),
  id       = "id",
  family   = "binomial"
)
spec
```

## 2. Run the whole pipeline in one call

`run_gbtm_pipeline()` performs algorithm selection (when the engine offers a
choice), group-number selection, the polynomial-shape search with GRoLTS
acceptance criteria, and the final Hessian-on fit. Here we use a small search to
keep the vignette quick.

```{r, eval = has_trajer}
res <- run_gbtm_pipeline(
  spec,
  candidates = 2:5,     # consider 2 to 5 groups
  degree     = 2,       # quadratic while choosing the number of groups --
                        # with curved shapes, linear-only selection under-selects
  method     = "L",     # fix the algorithm (skip stage 1) for speed
  max_degree = 3,       # allow up to cubic in the shape search
  seed       = 1,
  verbose    = FALSE
)
res
```

The pipeline recovers the four planted groups. Everything each stage produced is
kept on the result object:

```{r, eval = has_trajer}
res$group_selection      # BIC for each candidate number of groups
```

```{r, eval = has_trajer}
summary(res)
```

## 3. Inspect and plot

`gbtm_diagnostics()` gives the GRoLTS classification diagnostics, and
`plot_trajectories()` draws the fitted group trajectories with the observed
per-group means overlaid.

```{r, eval = has_trajer}
res$diagnostics$groups
```

```{r, eval = has_trajer && has_ggplot, fig.alt = "Fitted group trajectories"}
plot_trajectories(res$final_fit)
```

Per-subject group assignment (the analogue of exporting a group column) is in
`res$assignment`:

```{r, eval = has_trajer}
head(res$assignment)
```

## 4. Or run the stages individually

`run_gbtm_pipeline()` is a convenience wrapper. For full control you can run each
GRoLTS stage yourself and inspect the result before moving on -- this is exactly
what the wrapper does internally.

**Stage 1 -- choose the estimation algorithm.** trajeR offers `"L"`, `"EM"` and
`"EMIRLS"`; the one with the lowest BIC is selected. (Engines with a single
optimizer skip this stage.)

```{r, eval = has_trajer}
algo <- select_algorithm(spec, n_groups = 2, degrees = c(1, 1))
algo
```

**Stage 2 -- choose the number of groups** by BIC over a set of candidates.
Quadratic shapes are used while sweeping: with curved trajectories like these,
linear-only selection under-selects the number of groups:

```{r, eval = has_trajer}
groups <- select_n_groups(spec, candidates = 2:5, degree = 2, method = "L")
groups
```

**Stage 3 -- search polynomial shapes** for the chosen number of groups, then
apply the GRoLTS acceptance criteria (PMS > 0.05, APPA > 0.70, OCC >= 5). (The
one-call pipeline above searched up to cubic; here we cap the search at
quadratic to keep the vignette quick.)

```{r, eval = has_trajer}
shapes <- evaluate_shapes(spec, n_groups = groups$best, method = "L",
                          max_degree = 2, verbose = FALSE)
apply_grolts_criteria(shapes)
```

**Stage 4 -- fit the final model** (Hessian on, so the fit carries standard
errors) and read off the diagnostics and per-subject assignment. Here we use the
lowest-BIC shape found by the search:

```{r, eval = has_trajer}
fit <- fit_gbtm(spec, n_groups = groups$best, degrees = shapes$best, method = "L")
gbtm_diagnostics(fit)
head(gbtm_assign(fit))
```

## Continuous outcomes

The same pipeline handles continuous outcomes: switch the family to
`"gaussian"` (mapped to a censored-normal model) and point the spec at a
continuous dataset. `sim_continuous` has the same four shape types on a
continuous scale.

We fit cubic shapes in all four groups (in a real analysis the shape search
refines the per-group degrees, as above). Shape misspecification can push a fit
into a degenerate local optimum where a group ends up empty -- on this data,
forcing *linear* shapes (`degrees = rep(1, 4)`) with the `"L"` optimizer does
exactly that. If it happens, `plot_trajectories()` / `gbtm_predict()` warn you;
switching the method (`"EM"` recovers all four groups here) or revisiting the
shapes fixes it.

```{r, eval = has_trajer}
data("sim_continuous", package = "gbtmkit")
cspec <- gbtm_spec(
  sim_continuous,
  outcomes = paste0("y", 1:10),
  time     = paste0("t", 1:10),
  id       = "id",
  family   = "gaussian"
)
cfit <- fit_gbtm(cspec, n_groups = 4, degrees = rep(3, 4), method = "L")
gbtm_diagnostics(cfit)$entropy
```

`plot_trajectories()` works the same way; for a continuous outcome the fitted
lines and the observed points are on the outcome's own scale (means, not
probabilities):

```{r, eval = has_trajer && has_ggplot, fig.alt = "Fitted continuous group trajectories"}
plot_trajectories(cfit)
```

## Notes

* **Engine-agnostic.** The pipeline talks only to a small set of accessors
  (`gbtm_bic()`, `gbtm_posterior()`, `gbtm_group_sizes()`, ...), so additional
  backends can be added without changing the workflow.
* **Built to scale.** The shape search runs the fits with the Hessian off
  (needed only for the final model's standard errors), defaults to a greedy
  stepwise strategy, and supports a `time_budget`, `max_fits`, and on-disk
  `checkpoint`ing so large problems run unattended and bounded.
```
