---
title: "Getting started with gbtmkit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with gbtmkit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
# The estimation engine (trajeR) is an optional dependency; only run the fitting
# chunks when it is available so the vignette also builds without it.
has_trajer <- requireNamespace("trajeR", quietly = TRUE)
has_ggplot <- requireNamespace("ggplot2", quietly = TRUE)
```

`gbtmkit` turns group-based trajectory modeling (GBTM) into a reproducible
pipeline that follows the GRoLTS reporting checklist. Estimation is delegated to
interchangeable backends behind one interface, so the same workflow handles
binary and continuous outcomes, and the fit diagnostics (entropy, APPA, OCC,
group proportions) are computed the same way regardless of engine.

```{r setup}
library(gbtmkit)
```

## The data

The package ships two entirely synthetic datasets with known ground-truth
groups. `sim_binary` has a binary outcome measured on four occasions, with four
latent trajectory shapes (stable-high, falling, rising, stable-low):

```{r}
data("sim_binary", package = "gbtmkit")
head(sim_binary)
```

## 1. Describe the model with a spec

`gbtm_spec()` records *what* to model -- the outcome and time columns (by name),
the id, and the outcome family -- and validates it, independent of which engine
will fit it.

```{r}
spec <- gbtm_spec(
  sim_binary,
  outcomes = c("y1", "y2", "y3", "y4"),
  time     = c("t1", "t2", "t3", "t4"),
  id       = "id",
  family   = "binomial"
)
spec
```

## 2. Run the whole pipeline in one call

`run_gbtm_pipeline()` performs algorithm selection (when the engine offers a
choice), group-number selection, the polynomial-shape search with GRoLTS
acceptance criteria, and the final Hessian-on fit. Here we use a small search to
keep the vignette quick.

```{r, eval = has_trajer}
res <- run_gbtm_pipeline(
  spec,
  candidates = 2:5,     # consider 2 to 5 groups
  degree     = 1,       # linear shapes while choosing the number of groups
  method     = "L",     # fix the algorithm (skip stage 1) for speed
  max_degree = 2,       # allow up to quadratic in the shape search
  seed       = 1,
  verbose    = FALSE
)
res
```

The pipeline recovers the four planted groups. Everything each stage produced is
kept on the result object:

```{r, eval = has_trajer}
res$group_selection      # BIC for each candidate number of groups
```

```{r, eval = has_trajer}
summary(res)
```

## 3. Inspect and plot

`gbtm_diagnostics()` gives the GRoLTS classification diagnostics, and
`plot_trajectories()` draws the fitted group trajectories with the observed
per-group means overlaid.

```{r, eval = has_trajer}
res$diagnostics$groups
```

```{r, eval = has_trajer && has_ggplot, fig.alt = "Fitted group trajectories"}
plot_trajectories(res$final_fit)
```

Per-subject group assignment (the analogue of exporting a group column) is in
`res$assignment`:

```{r, eval = has_trajer}
head(res$assignment)
```

## Continuous outcomes

The same pipeline handles continuous outcomes: switch the family to
`"gaussian"` (mapped to a censored-normal model) and point the spec at a
continuous dataset. `sim_continuous` has the same four shape types on a
continuous scale.

```{r, eval = has_trajer}
data("sim_continuous", package = "gbtmkit")
cspec <- gbtm_spec(
  sim_continuous,
  outcomes = c("y1", "y2", "y3", "y4"),
  time     = c("t1", "t2", "t3", "t4"),
  id       = "id",
  family   = "gaussian"
)
cfit <- fit_gbtm(cspec, n_groups = 4, degrees = rep(1, 4), method = "L")
gbtm_diagnostics(cfit)$entropy
```

## Notes

* **Engine-agnostic.** The pipeline talks only to a small set of accessors
  (`gbtm_bic()`, `gbtm_posterior()`, `gbtm_group_sizes()`, ...), so additional
  backends can be added without changing the workflow.
* **Built to scale.** The shape search runs the fits with the Hessian off
  (needed only for the final model's standard errors), defaults to a greedy
  stepwise strategy, and supports a `time_budget`, `max_fits`, and on-disk
  `checkpoint`ing so large problems run unattended and bounded.
```
